
---
title: "Gapminder Assignment"
author: "Elle Wibisono, Maggie Tsai, Paul Carvalho, Xintong Guan"
date: "March 10, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=6, fig.height=3.5, warning=FALSE, message=FALSE)
```

## Analysis of gapminder dataset

Gapminder data includes population, life expectancy, and GDP per capita for 142 countries on 5 continents. Data was collected between years 1952 and 2007.

```{r gapminder, include=FALSE}
library(ggplot2)
library(plyr)
library(dplyr)
library(curl)
library(bbmle)

# Load the data
gapminder_location = curl(url = "https://raw.githubusercontent.com/resbaz/r-novice-gapminder-files/master/data/gapminder-FiveYearData.csv")
gapminder = read.csv(gapminder_location, stringsAsFactors = FALSE)

# Create a variable expressing population in millions.
gapminder$pop_mil = gapminder$pop/1000000

# Functions
  # Calculate standard error of the mean.
  se = function(x){
    std.error = sd(x)/sqrt(length(x))
    return(std.error)
  }
```

## Variables in relation to time and continent

```{r life expectancy, echo=FALSE}
# Split data by continent and year, then calculate the mean and standard error of the mean for life expectancy, population, and gdpPercap.
gap1 = ddply(gapminder, c('continent','year'), function(df)c(mean(df$lifeExp),
                                                             se(df$lifeExp),
                                                             mean(df$pop_mil),
                                                             se(df$pop_mil),
                                                             mean(df$gdpPercap),
                                                             se(df$gdpPercap)))
colnames(gap1)[3:8] = c('Mean.LifeExp',
                        'SE.LifeExp',
                        'Mean.PopMil',
                        'SE.PopMil',
                        'Mean.gdpPercap',
                        'SE.gdpPercap')

# Plot mean life expectency (including standard error of the mean) for each continent in relation to year.
ggplot(gap1, aes(year, Mean.LifeExp)) +
       geom_point(aes(color=continent)) +
       geom_line(aes(color=continent)) +
       geom_linerange(aes(ymin=Mean.LifeExp - SE.LifeExp,
                         ymax=Mean.LifeExp + SE.LifeExp,
                         color = continent)) +
       xlab("Year") +
       ylab("Life expectancy") +
       theme(legend.title=element_blank())+ ggtitle("Figure 1a. Mean Life Expectancy by Continent")

```

Figure 1a. Mean life expectancy by continent over time (years 1952-2007). Bars indicate standard error of the mean.

```{r life expectancy regression, echo=FALSE}
#Regression plot for life expectancy and year
#run regression to see p value and significance
ggplot(data=gapminder, aes(x=year, y=lifeExp))+geom_point()+geom_smooth(method=lm)+ ggtitle ("Figure 1b. Regression model for life expectancy by year")+ 
  xlab("Year") +
  ylab("Life Expectancies")
```
 
Figure 1b. Regression model for life expectancy by year. Life expectancy was significantly predicted by year p<0.001, with a large effect R2 = 0.19. 

```{r life expectancy regression summary stats, echo=FALSE, include=FALSE}

summary(lm(gapminder$lifeExp~gapminder$year)) 
```

```{r population, echo=FALSE}
# Plot mean population in millions (including standard error of the mean) for each continent in relation to year.
ggplot(gap1, aes(year, Mean.PopMil)) +
  geom_point(aes(color=continent)) +
  geom_line(aes(color=continent)) +
  geom_linerange(aes(ymin=Mean.PopMil - SE.PopMil,
                     ymax=Mean.PopMil + SE.PopMil,
                     color = continent)) +
  xlab("Year") +
  ylab("Population (millions)") +
  theme(legend.title=element_blank())
```

Figure 2a. Mean population by continent over time (years 1952-2007). Bars indicate standard error of the mean.

```{r population regression, echo=FALSE}
#run regression to see p value and significance
ggplot(data=gapminder, aes(x=year, y=pop))+geom_point()+geom_smooth(method=lm)+ ggtitle ("Figure 1b. Regression model for population by year")+
  xlab("Year") +
  ylab("Population (millions)")
```
Figure 2b. Regression model for population by year. Population in millions was significantly predicted by year p<0.001, with a minimal effect R2 = 0.006.

```{r population summary statistics, echo=FALSE, include=FALSE}
summary(lm(gapminder$pop~gapminder$year))
```

```{r gdp per cap, echo=FALSE}
# Plot mean gdp per capita (including standard error of the mean) for each continent in relation to year.
ggplot(gap1, aes(year, Mean.gdpPercap)) +
  geom_point(aes(color=continent)) +
  geom_line(aes(color=continent)) +
  geom_linerange(aes(ymin=Mean.gdpPercap - SE.gdpPercap,
                     ymax=Mean.gdpPercap + SE.gdpPercap,
                     color = continent)) +
  xlab("Year") +
  ylab("GDP per capita") +
  theme(legend.title=element_blank())+ggtitle("Figure 2. Mean Population by Continent")
```

Figure 3. Mean GDP per capita by continent over time (years 1952-2007). Bars indicate standard error of the mean.

## Life expectancy and GDP per capita

```{r lifeExp as a function of gdpPercap, echo=FALSE}
ggplot(gapminder, aes(log(gdpPercap), lifeExp)) +
       geom_point() +
       geom_smooth (method=lm, level=0.95) +
       xlab('GDP per capita (log)')+
       ylab('Life expectancy')+ ggtitle("Figure 3. Mean GDP per capita")
```

Figure 4. Life expectancy as a function of GDP per cap for all countries and years (1952-2007). The blue line indicates linear regression and shading indicates 95% confidence intervals.

Linear regression: lifeExp = [8.4 * log(gdpPercap)] - 9.1  
R-squared: 0.65

```{r lifeExp as a function of gdpPercap models, echo=FALSE, include=FALSE}
# Linear Models
fit1 = lm(gapminder$lifeExp ~ log(gapminder$gdpPercap))
summary(fit1)

fit2 = lm(gapminder$lifeExp ~ log(gapminder$gdpPercap) + gapminder$continent)
summary(fit2)

fit3 = lm(gapminder$lifeExp ~ log(gapminder$gdpPercap) + as.factor(gapminder$year))
summary(fit3)

fit4 = lm(gapminder$lifeExp ~ log(gapminder$gdpPercap) * as.factor(gapminder$year))
summary(fit4)

AICtab(fit1, fit2, fit3, fit4)

```
We created four different linear models and assessed fit with the Akaike Information criterion (AIC). The model with the best fit was an ANCOVA model which assigned the dependent variable as life expectancy, and the categorical independent variable as the year and the covariate as the GDP per capita. 
P- value= < 2.2e-16

```{r lifeExp as a function of gdpPercap, with year as a factor, echo=FALSE}
#Create plot based on the model fit 3
fit3anc <- predict(fit3)
ggplot(data=cbind(gapminder, fit3anc),
      aes(gapminder$lifeExp, log(gapminder$gdpPercap), color=year))+
      geom_point()+
      facet_wrap(~year, ncol = 3)+ 
      geom_smooth(method='lm', color= "lightgreen") + xlab('Life Expectancy') + 
      ylab('log GDP per Capita')
```
Figure 5. Life expectancy as a function of GDP per capita, with year as a factor. A line was fitted for each year. Based on the model chosen by the Akaike Information Criterion (AIC), GDP per capita is a predictor, with each subsequent year significantly different from that of year 1957. 
## Life Expectancy and Population

```{r lifeExp as a function of population in millions, echo=FALSE}

# Population and Life Expectancy 
ggplot(gap1, aes(log(Mean.PopMil), Mean.LifeExp)) +
  geom_point(aes(color=continent)) +
  geom_smooth (method=lm, level=0.95) +
  xlab('Mean Population in Millions (log)')+
  ylab('Mean Life expectancy')
```

```{r lifeExp as a function of population regression summary stats, echo=FALSE, include=FALSE}
#Life expectancy as a function of population 

LC1 = lm(gap1$Mean.LifeExp ~ log(gap1$Mean.PopMil))
summary(LC1)
#No significance between IV Population in Millions and DV Life Expectancy, p=0.389
```

Figure 6. Life expectancy as a function of population in millions. A linear model was fitted for the data. Population in millions did not significantly predict life expectancy p=0.389

## Life Expectancy and Continent

```{r life expectancy and continent boxplots, echo=FALSE}

# Life Expectancy and Continent boxplots

ggplot(gap1, aes(x = continent, y = Mean.LifeExp)) +
  geom_boxplot(fill = "skyblue2", colour = "dimgray") +
  scale_x_discrete() + xlab("Continent") +
  ylab("Mean Life Expectancy")
```

```{r life expectancy and continent, echo=FALSE, include=FALSE}
#One-Way Analysis of variance conducted to assess if life expectancies for all years significantly differed by continent

aov1 = aov(gap1$Mean.LifeExp ~ gap1$continent)
summary(aov1)

#Results show omnibus significance p<.001 with medium to large effect size of continent on life expectancy

#Post-hoc test multiple comparisons with Tukey HSD, alpha = 0.05

TukeyHSD(aov1)

#Non-significant contrasts: Asia-Americas, Oceania-Europe, alpha = 0.05
```
Figure 7. Boxplots for life expectancy per continent by the mean of all years and countries. A one-way analysis of variance was conducted, which revealed significant differences in life expectancy between continents p<001. Post-hoc TukeyHSD test showed that most comparisons of life expectancy between continents were significant. However, Asia did not significantly differ from America p=0.322, and Oceania did not significantly differ from Asia.  
