<<<<<<< HEAD
---
title: "Gapminder Assignment"
author: "Elle Wibisono, Maggie Tsai, Paul Carvalho"
date: "March 7, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=6, fig.height=3.5, warning=FALSE, message=FALSE)
```

## Analysis of gapminder dataset

Gapminder data includes population, life expectancy, and GDP per captita for 142 countries on 5 continents. Data was collected between years 1952 and 2007.

```{r gapminder, include=FALSE}
library(ggplot2)
library(plyr)
library(dplyr)
library(curl)
library(bbmle)

# Load the data
gapminder_location = curl(url = "https://raw.githubusercontent.com/resbaz/r-novice-gapminder-files/master/data/gapminder-FiveYearData.csv")
gapminder = read.csv(gapminder_location, stringsAsFactors = FALSE)

# Create a variable expressing population in millions.
gapminder$pop_mil = gapminder$pop/1000000

# Functions
  # Calculate standard error of the mean.
  se = function(x){
    std.error = sd(x)/sqrt(length(x))
    return(std.error)
  }
```

## Variables in relation to time and continent

```{r life expectancy, echo=FALSE}
# Split data by continent and year, then calculate the mean and standard error of the mean for life expectancy, population, and gdpPercap.
gap1 = ddply(gapminder, c('continent','year'), function(df)c(mean(df$lifeExp),
                                                             se(df$lifeExp),
                                                             mean(df$pop_mil),
                                                             se(df$pop_mil),
                                                             mean(df$gdpPercap),
                                                             se(df$gdpPercap)))
colnames(gap1)[3:8] = c('Mean.LifeExp',
                        'SE.LifeExp',
                        'Mean.PopMil',
                        'SE.PopMil',
                        'Mean.gdpPercap',
                        'SE.gdpPercap')

# Plot mean life expectency (including standard error of the mean) for each continent in relation to year.
ggplot(gap1, aes(year, Mean.LifeExp)) +
       geom_point(aes(color=continent)) +
       geom_line(aes(color=continent)) +
       geom_linerange(aes(ymin=Mean.LifeExp - SE.LifeExp,
                         ymax=Mean.LifeExp + SE.LifeExp,
                         color = continent)) +
       xlab("Year") +
       ylab("Life expectancy") +
       theme(legend.title=element_blank())
```

Figure 1. Mean life expectancy by continent over time (years 1952-2007). Bars indicate standard error of the mean.

```{r population, echo=FALSE}
# Plot mean population in millions (including standard error of the mean) for each continent in relation to year.
ggplot(gap1, aes(year, Mean.PopMil)) +
  geom_point(aes(color=continent)) +
  geom_line(aes(color=continent)) +
  geom_linerange(aes(ymin=Mean.PopMil - SE.PopMil,
                     ymax=Mean.PopMil + SE.PopMil,
                     color = continent)) +
  xlab("Year") +
  ylab("Population (millions)") +
  theme(legend.title=element_blank())
```

Figure 2. Mean population by continent over time (years 1952-2007). Bars indicate standard error of the mean.

```{r gdp per cap, echo=FALSE}
# Plot mean gdp per capita (including standard error of the mean) for each continent in relation to year.
ggplot(gap1, aes(year, Mean.gdpPercap)) +
  geom_point(aes(color=continent)) +
  geom_line(aes(color=continent)) +
  geom_linerange(aes(ymin=Mean.gdpPercap - SE.gdpPercap,
                     ymax=Mean.gdpPercap + SE.gdpPercap,
                     color = continent)) +
  xlab("Year") +
  ylab("GDP per capita") +
  theme(legend.title=element_blank())
```

Figure 3. Mean GDP per capita by continent over time (years 1952-2007). Bars indicate standard error of the mean.

## Life expectancy and GDP per capita

```{r lifeExp as a function of gdpPercap, echo=FALSE}
ggplot(gapminder, aes(log(gdpPercap), lifeExp)) +
       geom_point() +
       geom_smooth (method=lm, level=0.95) +
       xlab('GDP per capita (log)')+
       ylab('Life expectancy')
```

Figure 4. Life expectancy as a function of GDP per cap for all countries and years (1952-2007). The blue line indicates linear regression and shading indicates 95% confidence intervals.

Linear regression: lifeExp = [8.4 * log(gdpPercap)] - 9.1  
R-squared: 0.65

```{r lifeExp as a function of gdpPercap models, echo=FALSE, include=FALSE}
# Linear Models
fit1 = lm(gapminder$lifeExp ~ log(gapminder$gdpPercap))
summary(fit1)

fit2 = lm(gapminder$lifeExp ~ log(gapminder$gdpPercap) + gapminder$continent)
summary(fit2)

fit3 = lm(gapminder$lifeExp ~ log(gapminder$gdpPercap) + as.factor(gapminder$year))
summary(fit3)

fit4 = lm(gapminder$lifeExp ~ log(gapminder$gdpPercap) * as.factor(gapminder$year))
summary(fit4)

AICtab(fit1, fit2, fit3, fit4)

# Elle please plot linear model for fit3
```


=======
---
title: "Gapminder Assignment"
author: "Elle Wibisono, Maggie Tsai, Paul Carvalho"
date: "March 7, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=6, fig.height=3.5, warning=FALSE, message=FALSE)
```

## Analysis of gapminder dataset

Gapminder data includes population, life expectancy, and GDP per captita for 142 countries on 5 continents. Data was collected between years 1952 and 2007.

```{r gapminder, include=FALSE}
library(ggplot2)
library(plyr)
library(dplyr)
library(curl)
library(bbmle)

# Load the data
gapminder_location = curl(url = "https://raw.githubusercontent.com/resbaz/r-novice-gapminder-files/master/data/gapminder-FiveYearData.csv")
gapminder = read.csv(gapminder_location, stringsAsFactors = FALSE)

# Create a variable expressing population in millions.
gapminder$pop_mil = gapminder$pop/1000000

# Functions
  # Calculate standard error of the mean.
  se = function(x){
    std.error = sd(x)/sqrt(length(x))
    return(std.error)
  }
```

## Variables in relation to time and continent

```{r life expectancy, echo=FALSE}
# Split data by continent and year, then calculate the mean and standard error of the mean for life expectancy, population, and gdpPercap.
gap1 = ddply(gapminder, c('continent','year'), function(df)c(mean(df$lifeExp),
                                                             se(df$lifeExp),
                                                             mean(df$pop_mil),
                                                             se(df$pop_mil),
                                                             mean(df$gdpPercap),
                                                             se(df$gdpPercap)))
colnames(gap1)[3:8] = c('Mean.LifeExp',
                        'SE.LifeExp',
                        'Mean.PopMil',
                        'SE.PopMil',
                        'Mean.gdpPercap',
                        'SE.gdpPercap')

# Plot mean life expectency (including standard error of the mean) for each continent in relation to year.
ggplot(gap1, aes(year, Mean.LifeExp)) +
       geom_point(aes(color=continent)) +
       geom_line(aes(color=continent)) +
       geom_linerange(aes(ymin=Mean.LifeExp - SE.LifeExp,
                         ymax=Mean.LifeExp + SE.LifeExp,
                         color = continent)) +
       xlab("Year") +
       ylab("Life expectancy") +
       theme(legend.title=element_blank())
```

Figure 1. Mean life expectancy by continent over time (years 1952-2007). Bars indicate standard error of the mean.

```{r population, echo=FALSE}
# Plot mean population in millions (including standard error of the mean) for each continent in relation to year.
ggplot(gap1, aes(year, Mean.PopMil)) +
  geom_point(aes(color=continent)) +
  geom_line(aes(color=continent)) +
  geom_linerange(aes(ymin=Mean.PopMil - SE.PopMil,
                     ymax=Mean.PopMil + SE.PopMil,
                     color = continent)) +
  xlab("Year") +
  ylab("Population (millions)") +
  theme(legend.title=element_blank())
```

Figure 2. Mean population by continent over time (years 1952-2007). Bars indicate standard error of the mean.

```{r gdp per cap, echo=FALSE}
# Plot mean gdp per capita (including standard error of the mean) for each continent in relation to year.
ggplot(gap1, aes(year, Mean.gdpPercap)) +
  geom_point(aes(color=continent)) +
  geom_line(aes(color=continent)) +
  geom_linerange(aes(ymin=Mean.gdpPercap - SE.gdpPercap,
                     ymax=Mean.gdpPercap + SE.gdpPercap,
                     color = continent)) +
  xlab("Year") +
  ylab("GDP per capita") +
  theme(legend.title=element_blank())
```

Figure 3. Mean GDP per capita by continent over time (years 1952-2007). Bars indicate standard error of the mean.

## Life expectancy and GDP per capita

```{r lifeExp as a function of gdpPercap, echo=FALSE}
ggplot(gapminder, aes(log(gdpPercap), lifeExp)) +
       geom_point() +
       geom_smooth (method=lm, level=0.95) +
       xlab('GDP per capita (log)')+
       ylab('Life expectancy')
```

Figure 4. Life expectancy as a function of GDP per cap for all countries and years (1952-2007). The blue line indicates linear regression and shading indicates 95% confidence intervals.

Linear regression: lifeExp = [8.4 * log(gdpPercap)] - 9.1  
R-squared: 0.65

```{r lifeExp as a function of gdpPercap models, echo=FALSE, include=FALSE}
# Linear Models
fit1 = lm(gapminder$lifeExp ~ log(gapminder$gdpPercap))
summary(fit1)

fit2 = lm(gapminder$lifeExp ~ log(gapminder$gdpPercap) + gapminder$continent)
summary(fit2)

fit3 = lm(gapminder$lifeExp ~ log(gapminder$gdpPercap) + as.factor(gapminder$year))
summary(fit3)

fit4 = lm(gapminder$lifeExp ~ log(gapminder$gdpPercap) * as.factor(gapminder$year))
summary(fit4)

AICtab(fit1, fit2, fit3, fit4)

# Elle please plot linear model for fit3
```


#Xintong Guan add something to it

install.packages("ggplot2")
install.packages("plyr")
install.packages("dplyr")
install.packages("curl")
library(curl)

gapminder_location<-curl(url = "https://raw.githubusercontent.com/resbaz/r-novice-gapminder-files/master/data/gapminder-FiveYearData.csv")
gapminder<-read.csv(gapminder_location)

library("ggplot2")

#try all possible correlation between two of these variables, and draw figure of one of the model

ggplot(data=gapminder, aes(x=lifeExp, y=gdpPercap))+geom_point()+xlab("Life expectancy") + ylab("GDP per Capita")+ ggtitle("Figure 1")
ggplot(data=gapminder, aes(x=lifeExp, y=pop))+geom_point()
ggplot(data=gapminder, aes(x=lifeExp, y=year))+geom_point()
ggplot(data=gapminder, aes(x=pop, y=year))+geom_point()
ggplot(data=gapminder, aes(x=pop, y=continent))+geom_point()
ggplot(data=gapminder, aes(x=lifeExp, y=continent))+geom_point()
ggplot(data=gapminder, aes(x=gdpPercap, y=continent))+geom_point()

ggplot(data=gapminder, aes(x=year, y=lifeExp, color=continent))+geom_point()
ggplot(data=gapminder, aes(x=year, y=pop, color=continent))+geom_point()
ggplot(data=gapminder, aes(x=year, y=gdpPercap, color=continent))+geom_point()

#run regression of one model
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point() + scale_x_log10() + geom_smooth(method="lm")

gapminder.lm = lm(lifeExp ~ year, data=gapminder)
summary(gapminder.lm)

gdp<-gapminder$gdpPercap * gapminder$pop
gapminder <- cbind(gapminder, gdp)

ggplot(data=gapminder, aes(x=year, y=gdp, color=continent))+geom_point()
ggplot(data=gapminder, aes(x=lifeExp, y=gdp, color=continent))+geom_point()
ggplot(data=gapminder, aes(x=lifeExp, y=pop, color=continent))+geom_point()

ggplot(data=gapminder, aes(x=continent, y=gdp, color=year))+geom_point()

ggplot(data=gapminder, aes(x=lifeExp, y=gdp, color=year))+geom_point()
ggplot(data=gapminder, aes(x=lifeExp, y=pop, color=year))+geom_point()

#subset the data by country; 
#keep countries starting with "A" and "B"
#add captins and descriptions
subcountry <- substr(gapminder$country, start = 1, stop = 1)
AB.countries <- gapminder[subcountry %in% c("A", "B"), ]
ggplot(data = AB.countries, aes(x = year, y = lifeExp, color=continent)) +
  geom_line() + facet_wrap( ~ country) +
  xlab("Year") + ylab("Life expectancy") + ggtitle("Figure 2 Life Expectancy for A B country") +
  scale_colour_discrete(name="Continent")+theme(axis.text.x=element_blank(), axis.ticks.x=element_blank())

>>>>>>> a6aac1b4528cb4177b272984340ace96fe6bda3d
